---
title: "Installation"
format: html
---

Before you can begin on your new project, please complete the preparation steps relevant to your situation below. This will ensure that you will have all the tools/software needed for success! 

## 1️⃣ Create a GitHub Account

Create an account on [GitHub](https://github.com/) if you don't have one already. GitHub is a cloud-based platform and website where people can safely store and collaborate on code. You may have heard of it in relation to the software development world, but it is also huge in scientific research! Its useful capabilities will no doubt help elevate your workflow. 

Once you have your account, log in and make sure you're added to the [MDEL GitHub Organization](https://github.com/Malone-Disturbance-Ecology-Lab). All the relevant repositories belonging to the lab lives here. Once you're added, you will have permission to create and edit these repositories.

## 2️⃣ Install R / Python / Coding language of your choice

[Install R](https://www.r-project.org/) on your computer if you plan on using it. If you already have R and your R version is outdated by a few years, it is worth updating it at this point. Please ensure that your R version is at least 4.0.0 before proceeding to the next step.

[Install Python](https://www.python.org/downloads/) on your computer if you plan on using it. Again, if you already have Python, you might want to check to make sure you're not using a super outdated version.

Feel free to install any other coding languages as well.

## 3️⃣ Install RStudio / Visual Studio Code / IDE of your choice

Install [RStudio](https://docs.posit.co/ide/user/#direct-downloads-open-source) if you plan on using R. RStudio is currently the most popular integrated development environment (IDE) for R users, but it can also run Python.

You may also download other IDEs if you wish.

Another popular IDE is [Visual Studio Code](https://code.visualstudio.com/). This IDE is popular across lots of coding languages in general, and is a great choice if you plan on using various languages. 

This guide will have explicit tutorials geared towards RStudio and Visual Studio Code users, but even if you don't wish to use these, you may still find useful tips that can be applied to your own IDE of choice (see the "GitHub Desktop" section in step 7).

## 4️⃣ Check if you already have Git

If you're using a Mac computer, there's a good chance that it already has Git.

Run this command line code to check if your computer already has Git. You can run this in the Terminal on Mac or Linux. If you're using a Windows computer, and you don't know if you have an equivalent shell, you can open up RStudio or VS Code and run this command in the "Terminal" tab/window. 

```
which git
```

If this command returns a file path, then you already have Git and can skip the Git installation. Otherwise, proceed to install Git. 

## 5️⃣ Install Git

Git is an open-source version control software designed to track changes in files and allow other people to collaborate on your work. The way to install Git will be different depending on your operating system, so look through and pick the right option for your machine.

Be sure to close any IDEs you have open before installing Git, as this can cause your IDE to not be able to detect Git later on.

:::{.panel-tabset}
### Mac

Run the following command line code.

```
xcode-select --install
```

For more detailed instructions, see [here](https://happygitwithr.com/install-git#macos).

### Windows

[Install Git for Windows](https://gitforwindows.org/) (a.k.a. "Git Bash"). When asked about "Adjusting your PATH environment", **select "Git from the command line and also from 3rd-party software"**.

For more detailed instructions, see [here](https://happygitwithr.com/install-git#install-git-windows).

### Linux

Install Git via your distro's package manager (in the Terminal).

If you use Ubuntu or Debian Linux that code is as follows:

```
sudo apt-get install git
```

If instead you use Fedora or RedHat Linux the code is instead:

```
sudo yum install git
```

For more detailed instructions, see [here](https://happygitwithr.com/install-git#linux).

:::

## 6️⃣ Check that Git is installed

Run this command line code to check that your computer has Git. You can run this in the Terminal on Mac or Linux, or the Git Bash shell if you've installed Git for Windows from the previous step.

```
which git
```

## 7️⃣ Connect your IDE and Git/GitHub

In order to send code from your computer to GitHub, you will need to authenticate yourself so that GitHub knows you're actually the one making these changes. There are several ways to do this: the Personal Access Token (PAT), Secure Shell (SSH), or a browser-based OAuth authentication. We will focus on the PAT and OAuth methods for this walkthrough. For a more thorough breakdown of SSH, please see Jenny Bryan's "[HTTPS versus SSH](https://happygitwithr.com/https-pat#https-vs-ssh)" page.

Follow one of the RStudio or VS Code or GitHub Desktop tutorials for this step. You do not need to follow all 3. 

If you don't use RStudio or VS Code, try out GitHub Desktop instead!

:::{.panel-tabset}
### RStudio users

For this section, you can run the commands in the "**Console**" tab in RStudio.

First, make sure you have the `usethis` and `gitcreds` R packages installed. 

```{r}
#| eval: false
# Install the `usethis` and `gitcreds` packages
install.packages(c("usethis", "gitcreds"))
```

Then, update your Git profile on your computer. Replace `"Your GitHub Username"` with your actual GitHub username and replace `"Your GitHub Email"` with your actual email you use for GitHub. Once you filled those out, run the command.

```{r}
#| eval: false
# Add your GitHub username and email
usethis::use_git_config(user.name = "Your GitHub Username",
                        user.email = "Your GitHub Email")
```

Finally, create your Personal Access Token (PAT). You can run the command below to direct you to GitHub's page to create a new token. Once you're on that page, make sure to check the checkboxes to give yourself the necessary permissions. (You can just check everything if you're not sure).

When you get to the expiration date step, you may set it to 90 days, after which you will need to regenerate your token again. You will automatically get an email notification from GitHub when your token is approaching its expiration date.

For more help on PAT, check out steps 6-10 in [GitHub's PAT tutorial](https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens#creating-a-personal-access-token-classic).

```{r}
#| eval: false
# Create a token (Note this will open a GitHub browser tab)
## See steps 6-10 in GitHub's PAT tutorial 
usethis::create_github_token()
```

:::callout-note
This command will prompt you to create a "Classic" token on GitHub, but there are also a second type of token called "Fine-Grained" tokens. Fine-grained tokens will allow you to customize some very specific permissions, but Classic tokens will work just as well. 
:::

After generating your token, copy it. You may also want to paste it into a secure document to safely save it for future use. Head back to RStudio, type out and run this command. It will prompt you for your token. Paste it into the Console.

```{r}
#| eval: false
# Now, give your token to RStudio
## After you run this line follow the prompts in the "Console" tab of RStudio
gitcreds::gitcreds_set()
```

Now your computer will remember your token for the next 30, 60, 90, or so odd-days! Once your token expires, you will need to regenerate it and repeat this step. 

There you go! You should now be ready to do your workflow. Head to [Git Workflow](https://malone-disturbance-ecology-lab.github.io/Onboarding/workflow.html) for the next step.

### VS Code users

There are several ways you can connect your VS Code session and Git/GitHub through VS Code's handy interface! This guide will simply go over one possible method. For more options, feel free to read [VS Code's documentation](https://code.visualstudio.com/docs/sourcecontrol/github).

First, update your Git profile on your computer. You can do this by selecting **Terminal** --> **New Terminal**. Then, in the Terminal, run this command, but replace `"Your GitHub Username"` with your actual GitHub username. 

```
git config --global user.name "Your GitHub Username"
```
Next, run this command, but replace `"Your GitHub Email"` with your actual email you use for GitHub.

```
git config --global user.email "Your GitHub Email"
```
You can check your Git profile settings by running `git config -list` in the Terminal just to double-check.

Now click on your account menu in the bottom left corner of VS Code. Go to **Backup and Sync Settings...** --> **Sign in** --> **Sign in with GitHub**. 

A new window should have popped up to allow you to authorize VS Code. Follow the on-screen prompts to finish the browser-based OAuth authorization process.

Due to the way VS Code works, we will still need to finish the last of the authorization steps later on, so hang tight and proceed to [Git Workflow](https://malone-disturbance-ecology-lab.github.io/Onboarding/workflow.html) for the next step.

### GitHub Desktop

GitHub Desktop allows you to do the typical Git workflow, no matter what IDE you use. Since it is IDE-agnostic, it is extremely flexible in a variety of situations, even if your IDE does not have a dedicated GitHub feature.

Go [download GitHub Desktop](https://github.com/apps/desktop) and follow [Part 1 of their user guide](https://docs.github.com/en/desktop/overview/getting-started-with-github-desktop#part-1-installing-and-authenticating) to get started.

Then head to [Git Workflow](https://malone-disturbance-ecology-lab.github.io/Onboarding/workflow.html) for the next step.

:::